---
title: "Simulation experiment (1)"
format: 
  html:
    embed-resources: true
toc: true
---

```{r packages, echo = F}
library(targets)
library(tarchetypes)
library(tidyverse) |> suppressPackageStartupMessages()
library(here) |> suppressPackageStartupMessages()

Sys.setenv(TAR_PROJECT = "project_simexp1")
prj_store <- here("store_simexp1")
```

## Results

### Calibration metrics by scenario and model

```{r tar-load, echo = F}
tar_load(predeval_summary, store = prj_store)
nrep <- length(unique(predeval_summary$.rep)) 
tar_load(regdilution_summary, store = prj_store)
predeval_summary <- predeval_summary %>%
  arrange(sample_size, tau, tails, .rep) 
regdilution_summary <- regdilution_summary %>%
  arrange(sample_size, tau, tails)
regdilution_vctr <- rep(regdilution_summary$beta_obs, each = 3)
predeval_summary$regdilution <- regdilution_vctr
```

The three models were evaluated in 18 different scenarios. A scenario is a combination of the following parameters:

- Ratio between the SD of the measurement error and the SD of the latent $X$: $\sigma_u / \sigma_x$: 0.01, 0.1 or 0.2.
- Distribution of the measurement error: normal or student-t with 3 dof.
- Sample size of the dataset to train the transfer function: 100, 200, 500.

Each scenario was evaluated $S$ = `r nrep` times.

The following figures show the calibration metrics for all models per scenario-parameter. The combination of scenario-parameters is too cluttered to show in 1 Figure.

```{r, echo = FALSE}
#| label: fig-calibration1
#| fig-cap: "Results of the simulation experiment: relative size of the SD of the measurement error. Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(model, c("linreg","eivreg1","eivreg2"),
                   c("Linear regression","EIV model\n(known SD)",
                     "EIV model\n(prior on SD)"))
  ) %>%
  ggplot(aes(factor(tau), value, colour = model)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "",
    x = expression(sigma[u] / sigma[x]),
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, echo = FALSE}
#| label: fig-calibration2
#| fig-cap: "Results of the simulation experiment: sample size of the training dataset. Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(model, c("linreg","eivreg1","eivreg2"),
                   c("Linear regression","EIV model\n(known SD)",
                     "EIV model\n(prior on SD)"))
  ) %>%
  ggplot(aes(factor(sample_size), value, colour = model)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "",
    x = "sample size",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, echo = FALSE}
#| label: fig-calibration3
#| fig-cap: "Results of the simulation experiment: distribution of the measurement error. Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(model, c("linreg","eivreg1","eivreg2"),
                   c("Linear regression","EIV model\n(known SD)",
                     "EIV model\n(prior on SD)"))
  ) %>%
  ggplot(aes(tails, value, colour = model)) +
  facet_wrap(~ metric, scales = "free_y") +
  geom_boxplot() +
  geom_point(position = position_jitterdodge(.1, 0)) +
  labs(
    title = "calibration metrics", y = "",
    x = "distribution of the measurement error",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r, echo = FALSE, message=FALSE}
#| label: fig-calibration4
#| fig-cap: "Results of the simulation experiment: x-axis represents the amount of regression dilution in the training dataset (1 means no regression dilution, < 1 is more dilution, > 1 is regression 'concentration' which can happen in small samples by chance). Each facet has a different y-axis because the calibration metrics are on different scales."

predeval_summary %>%
  pivot_longer(
    c(PICP, MAE, MPE, RMSE, R2, SDPE), 
    names_to = "metric", values_to = "value"
  ) %>%
  mutate(
    metric = factor(metric, c("MPE","MAE","RMSE","R2","SDPE","PICP")),
    model = factor(model, c("linreg","eivreg1","eivreg2"),
                   c("Linear regression","EIV model\n(known SD)",
                     "EIV model\n(prior on SD)"))
  ) %>%
  ggplot(aes(regdilution, value, colour = model)) +
  facet_wrap(~ metric, scales = "free") +
  geom_point() + geom_smooth(method = "loess", se = F) +
  labs(
    title = "calibration metrics", y = "",
    x = "regression dilution",
    color = "Models"
  ) +
  theme_minimal() +
  theme(strip.text = element_text(face = "bold"),
        legend.position = "bottom")
```

### Model diagnostics: MCMC $\hat{R}$

The following figures display the $\hat{R}$ statistic of the MCMC for each of the model parameters. The red dashed lines are at 1.00 and 1.01. The boxplots are aggregated across scenario's because the values were similar in each of them and all well below 1.01.

```{r mcmc-dx, echo = F, warning = F}
tar_load(mcmcdx_linreg_all,  store = prj_store)
tar_load(mcmcdx_eivreg1_all, store = prj_store)
tar_load(mcmcdx_eivreg2_all, store = prj_store)

mcmcdx_linreg_all %>%
  select(-c(sample_size, tau, tails)) %>%
  pivot_longer(-(.rep), names_to = "parameter", values_to = "Rhat") %>%
  ggplot(aes(Rhat, parameter)) +
  geom_boxplot() +
  geom_vline(xintercept = c(1.00, 1.01), lty = 2, color = "red") +
  labs(title = "Linear model")

mcmcdx_eivreg1_all %>%
  select(-c(sample_size, tau, tails)) %>%
  pivot_longer(-(.rep), names_to = "parameter", values_to = "Rhat") %>%
  ggplot(aes(Rhat, parameter)) +
  geom_boxplot() +
  geom_vline(xintercept = c(1.00, 1.01), lty = 2, color = "red") +
  labs(title = "EIV model (known SD)")

mcmcdx_eivreg2_all %>%
  select(-c(sample_size, tau, tails)) %>%
  pivot_longer(-(.rep), names_to = "parameter", values_to = "Rhat") %>%
  ggplot(aes(Rhat, parameter)) +
  geom_boxplot() +
  geom_vline(xintercept = c(1.00, 1.01), lty = 2, color = "red") +
  labs(title = "EIV model (prior on SD)")
```

